Quantifying Limits to Detection of Early Warning for Critical Transitions: Supplementary Online Materials

### 2011-07-21 appendix/appendix.tex
minor edits from Alan

\documentclass[authoryear, preprint,review,12pt]{elsarticle}
%\usepackage{endfloat}
%\documentclass[final,5p,twocolumn]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{lineno}
\linenumbers
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{subfigure}
\usepackage[pdftex]{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.5,0}
%\usepackage[pdftex, colorlinks, citecolor=darkblue,linkcolor=darkgreen]{hyperref}
\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}

@@

\journal{Nature} 
\begin{document}
\begin{frontmatter}
\title{Quantifying Limits to Detection of Early Warning for Critical Transitions  }
\author[davis]{Carl Boettiger\corref{cor1}}
\ead{cboettig@ucdavis.edu}
\author[davis]{Alan M Hastings}
%\author[davis]{}
\cortext[cor1]{Corresponding author.}
\address[davis]{Center for Population Biology, University of California, Davis, United States}

%\begin{abstract}
%\end{abstract}

%\begin{keyword}
%\sep 
%\end{keyword}
\end{frontmatter}
%\include{figures.tex}
%input{figures.tex}
%\include{body.tex}
%\input{body.tex}

@@

\section{Method Summary}
For each dataset, parameters are estimated by maximum likelihood for a stable model and a model of a system approaching a critical transition (see Section~\ref{modelderivations}, and Section~\ref{likelihood} for the corresponding likelihood equations.)  The models are stochastic equations that can be used to simulate data under the same sampling regime as the original data, or at any specified set of time points.  Distributions are determined by a parametric bootstrap: 500 replicate datasets are simulated under both stable and deteriorating stability model (for each dataset).  For each replicate, the summary statistics (variance, autocorrelation, skew, and coefficient of variation) were computed over a sliding window equal to half the observed time, and a Kendall's correlation test applied to estimate correlation coefficient $\tau$.  The estimates of $\tau$ under the replicate simulations from the stable system and the estimates under the deteriorating stability system make up the two distributions which are integrated at different thresholds to create the ROC curves.

@@

\section{The data sets}\label{data}
All data sets and analyses performed are included in the accompanying R package.
The simulated data sets are produced with 40 data points
sampled evenly sampled over a time interval (0,100) using an individual-based model of a system containing a saddle-node bifurcation, see~\ref{modelderivations}.  The data was then fit to the LSN model (see~\ref{modelderivations}) to compute the ROC curves.  
%The deteriorating environment was produced under the LSN model~\eqref{LSN}, with $\theta=500$, $\sigma = 5$, $R(t) = 5 - 0.04999 t$.  
%The constant data is produced using the simulation method for the constant environment model~\eqref{OU},
%with parameters estimated from the deteriorating data-set to correspond as nearly as possible to a similar system.

The Glaciation data comes from~\citet{Petit1999}, accessible from NOAA:\\
\href{http://www.ncdc.noaa.gov/paleo/metadata/noaa-icecore-2453.html}{http://www.ncdc.noaa.gov/paleo/metadata/noaa-icecore-2453.html}.
The data is preprocessed by interpoplation and detrending to be as consistent as possible with the analysis presented in~\citet{Dakos2008},
analyzing the third glaciation event. 
The processed data consists of 121 sample points. 
The match is not exact since~\citet{Dakos2008} estimates the de-trending window size manually,
but the estimated correlations in the first-order auto-regression coefficients are in close agreement with that analysis. 
The code for processing the data from the original~\citet{Petit1999} file is included in the accompanying R package.

@@

where $f(x)$ and $g(x)$ are linear functions of $x$, the solution is Gaussian with moments evolving according the the coupled ordinary differential equations:

@@

These are numerically integrated using \texttt{lsoda} routine avialble in \texttt{R} for the likelihood calculation.

\section{The likelihood statistic}\label{Cox}
Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
which essentially asserts that comparing likelihoods is the most powerful test
of a choice between two hypotheses~\citep{Neyman1933}, which motivates
tests from the simple likelihood ratio test up through modern model adequacy methods.

@@

@@

The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''  %A statistical test is meaningless without at least some notion of power.  As it is perhaps the linguistics of hypothesis testing as much as the quantitative difficulty of the analysis that has led this step to be entirely ignored (though we admit some surprise at the extent to which the question of significance or false alarms has also been ignored), we hope

@@

\section{R package tutorial}
We provide an R package with simple implementation of the methods described here.  The package is available from: \href{https://github.com/cboettig/warningsignals/archives/master}{https://github.com/cboettig/warningsignals/archives/master}, where it will be actively maintained and developed.  The package takes an R time-series object (or, for unevenly spaced data -- a matrix or data-frame with sample times in the first column and observations in the second) as input and performs the likelihood-based analysis:

@@

The package also performs the computationally faster but less powerful summary-statistic tests, computes the resulting distributions under the different hypotheses and the corresponding ROC curves.  The package allows a variety of correlation tests (Pearson's, Kendall's, Spearman's) as the measure of increase, and a variety of summary statistics (variance, autocorrelation, skew, coeffient of variation).  Copies of data used in the paper are provided as examples.  The Monte Carlo methods can take advantage of multiple core processors or clusters by specifying the number of CPUs in the function argument.   Full documentation of functions and data are provided in the package, as well as example scripts which replicate the analyses presented in the paper.

\section{Examples}\label{examples}
In this section we illustrate the main results of the paper in three further empirical datasets from the climate record.

\pagebreak

>>>>

\documentclass[authoryear, preprint,review,12pt]{elsarticle}
%\usepackage{endfloat}
%\documentclass[final,5p,twocolumn]{elsarticle}
\bibliographystyle{elsarticle-harv}
\usepackage{lineno}
\linenumbers
\usepackage{graphicx}
\usepackage{amsmath,amsfonts}
\usepackage{subfigure, verbatim}
\usepackage[pdftex]{color}
\definecolor{darkblue}{rgb}{0,0,0.5}
\definecolor{darkgreen}{rgb}{0,0.5,0}
%\usepackage[pdftex, colorlinks, citecolor=darkblue,linkcolor=darkgreen]{hyperref}
\usepackage[pdftex, colorlinks]{hyperref}
\textwidth 6.75in
\oddsidemargin -0.15in
\evensidemargin -0.15in
\textheight 9in
\topmargin -0.5in
\newcommand{\ud}{\mathrm{d}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\C}{\mathrm{Cov}}
\newcommand{\V}{\mathrm{Var}}

@@

\journal{Science} 
\begin{document}
\begin{frontmatter}
\title{Quantifying Limits to Detection of Early Warning for Critical Transitions: Supplementary Online Materials}
\author[davis]{Carl Boettiger\corref{cor1}}
\ead{cboettig@ucdavis.edu}
\author[davis]{Alan M Hastings}
%\author[davis]{}
\cortext[cor1]{Corresponding author.}
\address[davis]{Center for Population Biology, University of California, Davis, United States}

\end{frontmatter}

@@

\section{Method Summary}
For each dataset, parameters are estimated by maximum likelihood for a stable model and a model of a system approaching a critical transition (see~\ref{modelderivations} and~\ref{likelihood} for the corresponding likelihood equations).  The models are stochastic equations that can be used to simulate data under the same sampling regime as the original data, or at any specified set of time points.  Distributions are determined by a parametric bootstrap: 500 replicate datasets are simulated under both stable and deteriorating stability model (for each dataset).  For each replicate, the summary statistics (variance, autocorrelation, skew, and coefficient of variation) were computed over a sliding window equal to half the observed time, and a Kendall's correlation test applied to estimate correlation coefficient $\tau$.  The estimates of $\tau$ under the replicate simulations from the stable system and the estimates under the deteriorating stability system make up the two distributions which are integrated at different thresholds to create the ROC curves.

@@

In addition to the descriptions provided here, the source code for implementing each of the analyses and generating the plots presented is provided as an R package,~\ref{R}. Further details on the implementation can be found in the function documentation.

\section{The data sets}\label{data}
All data sets and analyses performed are included in the accompanying R package, described in~\ref{R}.
The simulated data sets are produced with 40 data points
sampled evenly sampled over a time interval (0,100) using an individual-based model of a system containing a saddle-node bifurcation, see~\ref{modelderivations}.  The data was then fit to the LSN model, Eq~\eqref{LSN}, to compute the ROC curves. Code to run this individual-based simulation under a specified parameter values is provided in the \texttt{R} package.  See \verb|ibm_sims.R| demo for the example used here.  
%The deteriorating environment was produced under the LSN model~\eqref{LSN}, with $\theta=500$, $\sigma = 5$, $R(t) = 5 - 0.04999 t$.  
%The constant data is produced using the simulation method for the constant environment model~\eqref{OU},
%with parameters estimated from the deteriorating data-set to correspond as nearly as possible to a similar system.

The Glaciation data comes from~\citet{Petit1999}, accessible from NOAA:\\
\href{http://www.ncdc.noaa.gov/paleo/metadata/noaa-icecore-2453.html}{http://www.ncdc.noaa.gov/paleo/metadata/noaa-icecore-2453.html}.
The data is preprocessed by interpolation and de-trending to be as consistent as possible with the analysis presented in~\citet{Dakos2008},
analyzing the third glaciation event. 
The processed data consists of 121 sample points. 
The match is not exact since~\citet{Dakos2008} estimates the de-trending window size manually,
but the estimated correlations in the first-order auto-regression coefficients are in close agreement with that analysis. 
The code for processing the data from the original~\citet{Petit1999} file is included in the accompanying R package.

@@

where $f(x)$ and $g(x)$ are linear functions of $x$, the solution is Gaussian with moments evolving according the coupled ordinary differential equations:

@@

These are numerically integrated using \texttt{lsoda} routine available in \texttt{R} for the likelihood calculation.

\section{Comparing Models}\label{Cox}
Likelihood methods form the basis of much of modern statistics, in both Frequentist and Bayesian paradigms.  
The ability evaluate likelihoods directly by computation has made it possible to treat cases that do not conform to traditional assumptions more directly.
The basis of likelihood comparisons has its roots in the Neyman-Pearson Lemma, 
which essentially asserts that comparing likelihoods is the most powerful test
of a choice between two hypotheses~\citep{Neyman1933}, which motivates
tests from the simple likelihood ratio test up through modern model adequacy methods.

@@

\pagebreak
\section{The distributions for the ROC curves}
Figure 1 (main text) shows hypothetical distributions of a test statistic under a stable system and a system approaching a critical transition in order to illustrate how an ROC curve is calculated.  Figure~\ref{fig:3dist} shows the distributions used to generate the ROC curves in Figure 3 of the main text.

\begin{figure}[!ht]
  \begin{center}
    \includegraphics[width=.6\linewidth]{3dist}
  \end{center}
  \caption{The distributions underlying the ROC curves presented in Figure 3 of the main text.  As in Figure 1 (main text), the blue distribution corresponds to the stable model, while the red distribution corresponds to the model of a system approaching a critical transition.  Both models are estimated from the data by maximum likelihood.  The top panel shows the likelihood statistic, where the $x$-axis is the likelihood ratio $\delta$.  The increase in the various summary statistics is characterized by an Kendall's $\tau$.  The value of the statistic ($\tau$, $\delta$) estimated from the data set is shown by the triangle. Highly overlapping distributions correspond to ROC curves near the 1:1 line.}
  \label{fig:3dist}
\end{figure}

@@

The essential features of these overlapping distributions are easily captured in the ROC curves, which provides a natural language to discuss the issues of uncertainty -- such as ``accepting a higher false-alarm rate in order to avoid a catastrophic shift'' rather than ``using a more liberal threshold for $p$-values in order to avoid costly Type II error.''

@@

\section{R package tutorial}\label{R}
We provide an R package with simple implementation of the methods described here.  The package is available from: \href{https://github.com/cboettig/warningsignals/archives/master}{https://github.com/cboettig/warningsignals/archives/master}, where it will be actively maintained and developed.  The package takes an R time-series object (or, for unevenly spaced data -- a matrix or data-frame with sample times in the first column and observations in the second) as input and performs the likelihood-based analysis:

@@

The package also performs the computationally faster but less powerful summary-statistic tests, computes the resulting distributions under the different hypotheses and the corresponding ROC curves.  The package allows a variety of correlation tests (Pearson's, Kendall's, Spearman's) as the measure of increase, and a variety of summary statistics (variance, autocorrelation, skew, coefficient of variation).  Copies of data used in the paper are provided as examples.  The Monte Carlo methods can take advantage of multiple core processors or clusters by specifying the number of CPUs in the function argument.   Full documentation of functions and data are provided in the package, as well as example scripts which replicate the analyses presented in the paper.

\section{Examples}\label{examples}
In this section we illustrate the main results of the paper in several further empirical datasets from the climate record, Figs~\ref{fig:a3}, \ref{fig:a4}, and~\ref{fig:a3dist}.

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=\linewidth]{a3}
  \end{center}
  \caption{ROC curves for two additional data sets also analyzed in~\citet{Dakos2008}.  The area under each curve (AUC) is given in the legend. (a) The end of Greenhouse Earth period 34 million years ago may be detected in the 482 observations of CaCO$_3$ concentrations~\citep{Tripati2005}.   Despite the strong increase in autocorrelation suggested by the estimate of $\tau$, (see Fig~\ref{fig:a3dist}), the autocorrelation statistic performs little better than chance (AUC = 0.54). The variance in this signal actually decreases, causing the warning signal based on variance to perform substantially worse than chance.  Despite this, the likelihood approach still distinguishes reliably between the estimated deteriorating and stable hypotheses.  (b) Deuterium levels provide an early warning of the transition into the first ice age (Glaciation I) that appears in the time series from~\citep{Petit1999} (N=591).  Autocorrelation performs better than the other indicators, but again the likelihood approach is most reliable.  Figure~\ref{fig:a3dist} shows the corresponding distributions used to calculate the ROC curves.}
  \label{fig:a3}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=\linewidth]{a4}
  \end{center}
  \caption{ROC curves generated on sub-sampled data for the additional data sets.  The signal in the CaCO$_3$ data, Greenhouse Earth transition, is stronger than that in the deuterium data (Glaciation transition), making detection possible with fewer data points.}
  \label{fig:a4}
\end{figure}

\begin{figure}[ht]
  \begin{center}
    \includegraphics[width=\linewidth]{a3dist}
  \end{center}
  \caption{The distributions corresponding to the ROC curves in Fig~\ref{fig:a3}. The top panel shows the likelihood statistic, where the $x$-axis is the likelihood ratio $\delta$.  The increase in the various summary statistics is characterized by an Kendall's $\tau$.  The value of the statistic ($\tau$, $\delta$) estimated from the data set is shown by the triangle.   }
  \label{fig:a3dist}
\end{figure}
